---
title: "Portfolio Part 2 Plan"
author: "Dalton Garcia"

  
---

## Data Description

1)  My data source is from tidytuesdays and it is a set on Himalayan climbing expeditions.

2)  The three datasets consist of Himalayan peak, expedition, and members information. We will only be focusing on the peak and expedition datasets.

3)  My research questions that I hope to answer are:

  i)  What Himalayan peaks have the highest summit/survival rate? (Which mountains have been climbed the most with the least amount of deaths?)
  ii) What time of the year has the most successful summits? (No injuries/deaths)
  iii) What peak has the quickest average successful summits?

## Data Visualization

1)  
  i)  Interactive bar chart that displays sucessful summit rate and has the statistics of \# of summits and \# of deaths when hovered on.
  ii) Heatmap by month with a gradient color palette denoting an increase of deaths/accidents.
  iii) Perhaps 1 visualization of the quickest summits and then 1 visualization of the safest summits. \

2) I want to highlight the peaks that have the least amount of accidents and the time of year that has the least amount of accidents. I plan to do that by making my main focus of each visualization to be these issues.

3) I will need to do some data cleaning to filter certain outcomes of each expedition and create a number of new variables from the existing data in the set.

## Data Cleaning

3)  Do you need to filter your data? How? I will have to filter my expeditions by termination reason. For those expedition of success, I must filter this variable to only include those. This also applies for the expeditions where anyone got injured or died.

4)  Do you need to create any new variables? What variables? How? Yes, I need to create a duration variable. I will take the difference of the "highpoint_date"/termination_date" and basecamp_data" to find the duration of the trek.

5)  Do you need to add new data (join) to your data? What data? How? I must join my peak dataset with my expedition dataset. I will do this with a left join with the "peak_name" variable.


## Part 1- Data Cleaning
```{r}
library(tidyverse)
```


```{r}
#| label: load-in-data
#laoding in Tidy Tuesday Data
members <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-09-22/members.csv')
expeditions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-09-22/expeditions.csv')
peaks <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-09-22/peaks.csv')
#write.csv(members, "data-raw/members.csv")
#write.csv(expeditions, "data-raw/expeditions.csv")
#write.csv(peaks, "data-raw/peaks.csv")
```

```{r}
#| label: cleaning-data

glimpse(expeditions)
#added columns denoting if climb was successful or if any deaths occurred
expeditions <- expeditions |>
  mutate(climb_success = case_when(
    termination_reason == "Success (main peak)" ~ 1,
    termination_reason == "Success (subpeak)" ~ 1,
    termination_reason == "Success (claimed)" ~ 1,
    .default = 0),
          death_on_exp = if_else(member_deaths + hired_staff_deaths > 0, 1, 0))
            
#joined peaks and expeditions dataset to add height of peaks, dropped unused columns         
peak_expeditions <- expeditions |>
  left_join(peaks |> select(peak_name, height_metres), by = c('peak_name')) |>
  select(-oxygen_used, trekking_agency)


#successful summits creating length of trek to summit from basecamp
successful_expeditions <-  peak_expeditions |>
  filter(climb_success == 1) |>
  mutate(duration_to_summit = as.numeric(highpoint_date - basecamp_date))
  
```


```{r}
#| label: bar-plot-summit-rate
#summit rate depends on things like weather, running out of food, turning back for various reasons, aka not reaching the top
peak_expeditions |>
  group_by(peak_name) |>
  summarize(total_expeditions = n(),
            successful_treks = sum(climb_success),
            death_on_trek = sum(death_on_exp)) |>
  mutate(summit_rate = successful_treks/total_expeditions,
         death_rate = death_on_trek/total_expeditions) |>
  filter(total_expeditions >= 50) |>
  arrange(desc(total_expeditions)) |>
  head(10) |>
  
  #top 10 peaks by total # of expeditions
  ggplot(aes(x = fct_reorder(peak_name, total_expeditions, .desc = TRUE), y = total_expeditions, fill = peak_name)) +
  geom_col() +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Most Popular Himalayan Peaks by Total Expeditions (1905-2019)",
    x = "Peak Name",
    y = "Number of Expeditions"
  )  +
  theme_bw() +
  theme(legend.position = "none")

```

